---
title: "RNA-seq analysis of FL125 FL145 BM2W BM8W"
author: "heping"
date: "2026-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# I. load packages
```{r}
library(tidyverse)
library(ggplot2)
library(clusterProfiler)
library(ggpubr)
library(readxl)
library(ggrepel)
library(scales)
library(DESeq2)
library(stringr)
library(RColorBrewer)
library(paletteer)
library(pheatmap)
library(org.Mm.eg.db)
library(GSVAdata)
library(GSVA)
```

# II. load data
```{r}
counts <- read.csv("./FLBM.raw.count.matrix.csv") %>% column_to_rownames(var = "X")
tpm <- read.csv("./FLBM.tpm.csv")%>% column_to_rownames(var = "X")
colnames(tpm) <- c("FL12.5_1", "FL12.5_2", "FL12.5_3", "FL12.5_4",
                   "FL14.5_1", "FL14.5_2", "FL14.5_3", "FL14.5_4",
                   "BM2W_1", "BM2W_2", "BM2W_3", "BM2W_4",
                   "BM8W_1", "BM8W_2", "BM8W_3", "BM8W_4")
tpm <-tpm[rowMeans(tpm)>0.1,]
```

# III. colors
```{r}
group_color=c(FL12.5 = "#1B9E77", FL14.5 = "#2d6831",
              BM2W = "#f0acbd", BM8W = "#eb6b88")


coldata <- data.frame(Samples=colnames(data),
                      Group=c(rep("FL125",3),rep("FL145",3), rep("BM2W",3), rep("BM8W", 3)),
                      replicate=c(rep(1:3, 4)))
coldata$Group <- factor(coldata$Group, levels = c("FL12.5", "FL14.5", "BM2W", "BM8W"))
```

#=============================== Figure S2A-C ===============================
## Differential expressed genes
```{r}
coldata <- data.frame(Samples=colnames(data),
                      Group=c(rep("FL12.5",4),rep("FL14.5",4), rep("BM2W",4), rep("BM8W", 4)),
                      replicate=c(rep(1:4, 4)))
coldata$Group <- factor(coldata$Group, levels = c("FL12.5", "FL14.5", "BM2W", "BM8W"))

dds <- DESeqDataSetFromMatrix(counts, colData=coldata, design= ~ replicate+Group)
dds <- DESeq(dds)
```

## Figure S2A: PCA plot
```{r}
vsd <- vst(dds, blind=FALSE)
data <- assay(vsd)
data <- limma::removeBatchEffect(data, vsd$replicate)
data <- t(as.matrix(data))
pca.res <- prcomp(data, scale=T)

percentage<-round(pca.res$sdev / sum(pca.res$sdev) * 100,2)
percentage<-paste(colnames(pca.res),"(", paste(as.character(percentage), "%", ")", sep=""))
as.data.frame(pca.res$x) %>% rownames_to_column(var = "sample") %>%
  mutate(Treat=c(rep("FL12.5",4),rep("FL14.5",4),rep("BM2W",4),rep("BM8W",4))) %>%
  ggplot(aes(x=PC1,y=PC2,color=Treat,label=sample)) + 
  geom_point(size=3)+
  geom_text_repel()+
  scale_color_manual(values = group_color)+
  xlab(paste0("PC1",percentage[1])) + ylab(paste0("PC2",percentage[2])) + 
  theme_bw()
```

## Figure S2B: compare to last timepoint
```{r}
# FL145 vs FL125
diff_FL145_FL125 <- results(dds, contrast=c("Group","FL14.5","FL12.5"))
diff_FL145_FL125 <- as.data.frame(diff_FL145_FL125) %>% 
  filter(log2FoldChange != 'NA') %>%
  filter(padj != 'NA')
diff_FL145_FL125$type <- if_else(diff_FL145_FL125$padj<0.05 & diff_FL145_FL125$log2FoldChange>1.5,"up",
                        if_else(diff_FL145_FL125$padj<0.05 & diff_FL145_FL125$log2FoldChange<(-1.5),"down","noSig"))
# BM2W vs FL145
diff_BM2W_FL145 <- results(dds, contrast=c("Group","BM2W","FL14.5"))
diff_BM2W_FL145 <- as.data.frame(diff_BM2W_FL145) %>% 
  filter(log2FoldChange != 'NA') %>%
  filter(padj != 'NA')
diff_BM2W_FL145$type <- if_else(diff_BM2W_FL145$padj<0.05 & diff_BM2W_FL145$log2FoldChange>1.5,"up",
                        if_else(diff_BM2W_FL145$padj<0.05 & diff_BM2W_FL145$log2FoldChange<(-1.5),"down","noSig"))
# BM8W vs BM2W
diff_BM8W_BM2W <- results(dds, contrast=c("Group","BM8W","BM2W"))
diff_BM8W_BM2W <- as.data.frame(diff_BM8W_BM2W) %>% 
  filter(log2FoldChange != 'NA') %>%
  filter(padj != 'NA')
diff_BM8W_BM2W$type <- if_else(diff_BM8W_BM2W$padj<0.05 & diff_BM8W_BM2W$log2FoldChange>1.5,"up",
                                if_else(diff_BM8W_BM2W$padj<0.05 & diff_BM8W_BM2W$log2FoldChange<(-1.5),"down","noSig"))

# stats of DEGs
diff.list <- list(diff_FL145_FL125 = diff_FL145_FL125, diff_BM2W_FL145 = diff_BM2W_FL145,
                  diff_BM8W_BM2W = diff_BM8W_BM2W)
diff.gene.list <- lapply(1:length(diff.list), function(i, x){
  Up.gene <- x[[i]] %>% filter(type == "up") %>% pull(X)
  Up.num <- length(Up.gene)
  Down.gene <- x[[i]] %>% filter(type == "down") %>% pull(X)
  Down.num <- length(Down.gene)
  return(list(Up.num=Up.num, Down.num = Down.num))
}, x=diff.list)
names(diff.gene.list) <- names(diff.list)

diff.num.stats <- Reduce(rbind, diff.gene.list)
rownames(diff.num.stats) <- names(diff.gene.list)

data.plot <- diff.num.stats %>% as.data.frame() %>% 
  rownames_to_column(var = "group") %>% 
  pivot_longer(2:3, names_to = "type", values_to = "num") 

data.plot$type <- factor(data.plot$type, levels = c("Up.num","Down.num"))
data.plot$group <- factor(data.plot$group, levels = names(diff.gene.list))
data.plot$num <- as.numeric(data.plot$num)
ggbarplot(data.plot, x="group", y="num", fill = "type", 
          palette = c("#EC7357","#84B1ED"),position = position_dodge(width = 0.8),
          label = T)
```

## Figure S2C: pathway enrichment analysis
```{r}
diff.gene.list <- lapply(1:length(diff.list), function(i, x){
  Up.gene <- x[[i]] %>% filter(type == "up") %>% pull(X)
  Down.gene <- x[[i]] %>% filter(type == "down") %>% pull(X)
  return(list(Up.gene=Up.gene, Down.gene = Down.gene))
}, x=diff.list)
names(diff.gene.list) <- names(diff.list)

diff.gene <- list(up.FL145.FL125=diff.gene.list$diff_FL145_FL125$Up.gene,
                  down.FL145.FL125=diff.gene.list$diff_FL145_FL125$Down.gene,
                  up.BM2W.FL145=diff.gene.list$diff_BM2W_FL145$Up.gene,
                  down.BM2W.FL145=diff.gene.list$diff_BM2W_FL145$Down.gene,
                  up.BM8W.BM2W=diff.gene.list$diff_BM8W_BM2W$Up.gene,
                  down.BM8W.BM2W=diff.gene.list$diff_BM8W_BM2W$Down.gene)

lapply(diff.gene, function(x){tmp = bitr(x,
                                         fromType="SYMBOL", 
                                         toType="ENTREZID",
                                         OrgDb="org.Mm.eg.db")
return(tmp$ENTREZID)})->age.ensenble

go.res <- compareCluster(age.ensenble, fun = "enrichGO",
                         OrgDb='org.Mm.eg.db',ont="BP")
dotplot(go.res)+theme(axis.text.x = element_text(angle=60, hjust = 1, vjust = 1))
```

#=============================== Figure 1B,D-G & Figure S3A ===============================
## Figure 1B: Clustering analysis
```{r}
tpm <- log2(tpm+1)

mean <- tpm %>% rownames_to_column(var = "gene") %>% 
  pivot_longer(cols = 2:17, values_to = "expr", names_to = "sample")
mean$Group <- str_split(mean$sample, "\\_", simplify = T)[,1]
mean1 <- mean %>% group_by(gene, Group) %>% summarise(mean=mean(expr))
expr <- pivot_wider(mean1, names_from = "Group", values_from = "mean")
expr <- column_to_rownames(expr, var = "gene")
expr <- dplyr::select(expr, FL12.5, FL14.5, BM2W, BM8W)

getClusters(exp = expr)->p # chose cluster.num = 4
ck <- clusterData(exp = expr, cluster.method = "mfuzz", cluster.num = 4)
visCluster(object = ck, plot.type = "line")->p

if(T){
  pdf('./figure/heatmap.cluster.n4.pdf',height = 7,width = 7)
  visCluster(object = ck, plot.type = "both", column_names_rot = 360,
             cluster.order = c(3,1,4,2),
             show_row_dend = F,
             add.box = T, add.line = T, boxcol = group_color,
             line.side = "right")
  dev.off()
}
```

## Figure 1D-E & Figure S3A: GSVA analysis
```{r}
# GO_BP GSVA
my_bp<-getGmt("./data/m5.go.bp.v2023.2.Mm.symbols.gmt",
              geneIdType=SymbolIdentifier(),
              collectionType=BroadCollection(category="c5"), 
              sep="\t")

tpm.list <- list(tpm = tpm)
go.bp.list<- lapply(1:length(tpm.list), function(x,i){enrich_fun(x[[i]], my_bp, "logtpm")},
                    x=tpm.list)
names(go.bp.list) <- names(tpm.list)
go.bp.res <- go.bp.list$tpm

enrich_fun <- function(x,my_bp,data=c("logtpm", "count")){
  if(data=="logtpm"){
    ssgseaParam(exprData = as.matrix(log2(x+1)),
                geneSets=my_bp, 
                minSize=2, 
                maxSize=500)->res
    res <- gsva(res)
    return(res)
  }
  if(data=="count"){
    gsva(as.matrix(x), 
         my_bp, 
         min.sz=10, method="ssgsea",
         max.sz=500,kcdf="Poisson", 
         mx.diff=TRUE, verbose=FALSE, 
         parallel.sz=1, ssgsea.norm=TRUE)->res
    return(res)
  }
}

cell.path <- dplyr::filter(as.data.frame(go.bp.res), 
                           rownames(go.bp.res) %in%
                             c("GOBP_INNATE_IMMUNE_RESPONSE_ACTIVATING_SIGNALING_PATHWAY","GOBP_REGULATION_OF_INNATE_IMMUNE_RESPONSE",
                               "GOBP_ANTIVIRAL_INNATE_IMMUNE_RESPONSE","GOBP_INTERFERON_MEDIATED_SIGNALING_PATHWAY"))

cell.path <- dplyr::filter(as.data.frame(go.bp.res), 
                           rownames(go.bp.res) %in%
                             c("GOBP_ACTIVATION_OF_INNATE_IMMUNE_RESPONSE"))

cell.path <- cell.path %>% rownames_to_column("path")
colnames(cell.path) <- c("path",rep(paste0("FL12.5_",1:4)), rep(paste0("FL14.5_",1:4)),
                         rep(paste0("BM2W_",1:4)), rep(paste0("BM8W_",1:4)))

data <- cell.path %>%
  pivot_longer(cols = 2:17, values_to = "cell.score", names_to = "sample") %>%
  mutate(Group=str_split(sample, "\\_", simplify = T)[,1])
data$path <- str_remove(data$path, "GOBP_")
data$path <- str_to_title(data$path)
data$Group <- factor(data$Group, levels = names(group_color))


my_comparisons <- list(c("FL12.5","FL14.5"), c("FL14.5","BM2W"), c("BM2W","BM8W"))
ggboxplot(data, x="Group", y="cell.score", 
          fill = "Group")+facet_wrap(facets = "path")+
  scale_fill_manual(values = group_color)+
  stat_compare_means(comparisons=my_comparisons,label="p.signif")+
  theme_bw()+
  #ggtitle("GOBP Activation of innate immune response")+
  xlab("")+ylab("GSVA score")
```

```{r}
# Innate immune score
im.gene <- read.table("./data/innateDB_gene_lists_mm.txt") %>% pull(V1) #downloaded from innateDB
pathways <- list(im.gene=im.gene)

library(GSVA)
tmp.score <- gsva(as.matrix(log2(tpm+1)), 
                  pathways, min.sz=0, max.sz=1000,
                  kcdf="Gaussian", mx.diff=TRUE, verbose=FALSE,
                  parallel.sz=4, ssgsea.norm=T)

cell.path <- as.data.frame(tmp.score) %>% rownames_to_column("path")
colnames(cell.path) <- c("path",rep(paste0("FL12.5_",1:4)), rep(paste0("FL14.5_",1:4)),
                         rep(paste0("BM2W_",1:4)), rep(paste0("BM8W_",1:4)))

data <- cell.path %>%
  pivot_longer(cols = 2:17, values_to = "cell.score", names_to = "sample") %>%
  mutate(Group=str_split(sample, "\\_", simplify = T)[,1])
data$path <- str_remove(data$path, "GOBP_")
data$path <- str_to_title(data$path)
data$Group <- factor(data$Group, levels = names(group_color))

my_comparisons <- list(c("FL12.5","FL14.5"), c("FL14.5","BM2W"), c("BM2W","BM8W"), c("FL12.5","BM8W"))
ggboxplot(data, x="Group", y="cell.score", fill = "Group")+
  scale_fill_manual(values = group_color)+
  stat_compare_means(comparisons=my_comparisons,label="p.signif")+
  theme_bw()+
  ggtitle("Innte immune genes (InnateDB)")+
  xlab("")+ylab("GSVA score")
```

#=============================== Figure 1F: Heatmap of innate immune genes and globins ===============================
```{r}
select <- c("Lgals9","Oas1b","Oas1c","Oas3","Oas2","Isg15","Cd36","Ms4a8a",
            "Txnip","Bst2","Ccr3","Cxcr4",
            "Hbb-y","Hbb-bh1","Hbb-b1")
if(T){
  data<-tpm[select,]
  annotation_col = data.frame(CellType = factor(c(rep("FL12.5",4) ,rep("FL14.5",4),
                                                  rep("BM2W",4), rep("BM8W",4))))
  ann_colors = list( CellType = c(FL12.5 = "#1B9E77", FL14.5 = "#2d6831",
                                  BM2W = "#f0acbd", BM8W = "#eb6b88"))
  rownames(annotation_col) = colnames(data)
  data <- na.omit(data)
  library(pheatmap)
  pheatmap::pheatmap(data, scale = "row",
                     cluster_rows = F,cluster_cols = F, border_color = "white",
                     cellwidth = 10, cellheight = 10, 
                     annotation_col = annotation_col,
                     gaps_row = c(12, 15), 
                     gaps_col = 8,
                     show_colnames=F, annotation_colors = ann_colors)
}
```

#=============================== Figure 1G: Correlation of innate immune score and globins ===============================
```{r}
expr <- tpm[c("Hbb-b1"), ]
expr <- tpm[c("Hbb-y","Hbb-bh1"), ]

expr <- log2(expr+1)
expr <- as.data.frame(t(scale(t(expr))))
data <- rbind(tmp.score, expr)
colnames(data) <- c(rep(paste0("FL12.5_",1:4)), rep(paste0("FL14.5_",1:4)),
                    rep(paste0("BM2W_",1:4)), rep(paste0("BM8W_",1:4)))

tmp.plot <- t(data) %>% as.data.frame()
tmp.plot$avg <- (tmp.plot$`Hbb-b1`)
tmp.plot$avg <- (tmp.plot$`Hbb-y`+tmp.plot$`Hbb-bh1`)/2

tmp.plot <- tmp.plot %>% rownames_to_column(var="sample")
tmp.plot$sample <- str_split(tmp.plot$sample, "\\_", simplify = T)[,1]
tmp.plot$sample <- factor(tmp.plot$sample, levels = names(group_color))

cor.test(tmp.plot$im.gene, tmp.plot$avg, method = "spearman")

ggplot(tmp.plot, aes(x=im.gene, y=`avg`)) +
  geom_point(aes(color=sample), size=2.5) +
  #geom_smooth(method="lm", color="gray",fill = "lightgray") +
  geom_smooth(method="loess", color="gray",fill = "lightgray") +
  #labs(x = "Innate immune score", y = "Embryonic globin expression (Hbb-y, Hbb-bh1)") +
  labs(x = "Innate immune score", y = "Adult globin expression (Hbb-b1)") +
  scale_color_manual(values = group_color)+
  theme_bw()
```

